Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-28 paper.tex
Move bibliography to separate page
Apply languagetool suggestions

% TODO: check over/underfull boxes
% TODO: move references to new page

@@

ACL2, PVS and Aesop additionally provide special support for \emph{forward reasoning}, also known as \emph{forward chaining}, in the form of \emph{forward rules}.
Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
Hence forward rules can be used to establish additional facts that may be relevant for other rules.
For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules in its proof automation to establish facts that are later used by an arithmetic decision procedure.

@@

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.

@@

  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(m)$, the substitutions $\sub_{i}(m(i))$ and $\sub_{j}(m(j))$ are \emph{compatible}.
        This means that the substitutions agree on all common variables, i.e.\ for each variable $x ∈ \dom(\sub_{i}(m(i))) ∩ \dom(\sub_{j}(m(j)))$ we have $\sub_{i}(m(i))(x) = \sub_{j}(m(j))(x)$.
\end{enumerate}
The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
This is well defined by requirement~\ref{cond:match:compat}.
The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.

A partial match $m$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.

Note that matches represent only those partial applications where arguments are added without gaps.
Hence for a rule $r : A → B → C$, the applications $r~a~\_ : B \to C$ and $r~a~b : C$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to C$.
This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi have a similar effect.

@@

Given a hypothesis $h : T$ that matches slot $i > 1$ with substitution $σ = \sub_{i}(h)$, the partial matches that can be extended by $h$ are those in slot $i - 1$ whose substitutions are compatible with $σ$ on the variables $\vars(i)$ (i.e., the variables shared between $i$ and any earlier slot).
Assuming for the moment that $\vars(i)$ is nonempty for all slots $i > 1$, these matches are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))
\]
Similarly, given a partial match $m$ at level $i$ with substitution $σ = \sub(m)$, the hypotheses that could extend $m$ are
\[
  \bigcap_{x \in \vars(i+1)} H_{x}(i+1, σ(x))
\]
The number of shared variables tends to be small in practice, so these intersections can be computed efficiently.

@@

The rule state's $\addMatch$ method (Algorithm~\ref{alg:addmatch}) adds a match $m$ with level $i = \lvl(m)$ to the rule state.
If $m$ is already complete, it is returned.
Otherwise $m$ is inserted into $M$ and we query $H$ for hypotheses that can extend $m$.
Any extended matches thus produced are added to the rule state recursively.

\para{Variable Clusters.}
We have so far only considered rules where for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
However, there are also rules such as $\All{x~y}{P~x → Q~y → R~x~y}$, where slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
In this case, any match at level 1 can be extended with any hypothesis at level 2, since the two slots are independent.

@@

Now we add $h₂ : a ≤ b$, which again unifies with both slots.
For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $H_{y}(2, b) = \{h₁\}$.
Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
For slot~2, we add $h₂$ to $H_{y}(2, a)$.
There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.

@@

To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
Hence the compatible hypotheses are
\[
  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
\]
This means the match cannot currently be extended.
We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
This yields the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1, h_3\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}}
      child {node {$b$}
        child {node {$\{h_3\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$a$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$b$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}
Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
\]
This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.

@@

With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs are syntactically equal except for the marked proof subexpressions.
This notion of equality is syntactic, so it can be checked quickly and we can define an effective hash function that is compatible with it.
This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.

@@

\para{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.

@@

\para{Lazy Insertion.}
Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
We therefore delay insertions in two ways.
First, we observe that complete matches can only be generated once all slots have been filled.
Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found a potentially matching hypothesis for each slot.
Second, we observe that the variable indices are only queried for hypotheses at a slot $i > 1$ when a match is added at slot $i - 1$.
Hence we delay the insertion of hypotheses at $i$ until there is at least one match in slot $i - 1$.

@@

For each benchmark, we set up certain forward rules and a goal with handpicked hypotheses.
We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
Performing only forward reasoning is the best-case scenario for the incremental algorithm since previous results can be fully reused.
When forward reasoning is combined with other rules, we would expect smaller gains, depending on how much reuse the non-forward rules allow.

@@

\bibliography{lit}

>>>>

@@

ACL2, PVS and Aesop additionally provide special support for \emph{forward reasoning}, also known as \emph{forward chaining}, in the form of \emph{forward rules}.
Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
Hence, forward rules can be used to establish additional facts that may be relevant for other rules.
For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules in its proof automation to establish facts that are later used by an arithmetic decision procedure.

@@

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
Hence, it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.

@@

  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(m)$, the substitutions $\sub_{i}(m(i))$ and $\sub_{j}(m(j))$ are \emph{compatible}.
        This means that the substitutions agree on all common variables, i.e.\ for each variable $x ∈ \dom(\sub_{i}(m(i))) ∩ \dom(\sub_{j}(m(j)))$ we have $\sub_{i}(m(i))(x) = \sub_{j}(m(j))(x)$.
\end{enumerate}
The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
This is well-defined by requirement~\ref{cond:match:compat}.
The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.

A partial match $m$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$, and we have a full application.

Note that matches represent only those partial applications where arguments are added without gaps.
Hence, for a rule $r : A → B → C$, the applications $r~a~\_ : B \to C$ and $r~a~b : C$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to C$.
This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi have a similar effect.

@@

Given a hypothesis $h : T$ that matches slot $i > 1$ with substitution $σ = \sub_{i}(h)$, the partial matches that can be extended by $h$ are those in slot $i - 1$ whose substitutions are compatible with $σ$ on the variables $\vars(i)$ (i.e., the variables shared between $i$ and any earlier slot).
Assuming for the moment that $\vars(i)$ is non-empty for all slots $i > 1$, these matches are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))
\]
Similarly, given a partial match $m$ at level $i$ with substitution $σ = \sub(m)$, the hypotheses that could extend $m$ are
\[
  \bigcap_{x \in \vars(i+1)} H_{x}(i+1, σ(x))
\]
The number of shared variables tends to be small in practice, so these intersections can be computed efficiently.

@@

The rule state's $\addMatch$ method (Algorithm~\ref{alg:addmatch}) adds a match $m$ with level $i = \lvl(m)$ to the rule state.
If $m$ is already complete, it is returned.
Otherwise, $m$ is inserted into $M$ and we query $H$ for hypotheses that can extend $m$.
Any extended matches thus produced are added to the rule state recursively.

\para{Variable Clusters.}
We have so far only considered rules where for each slot $i > 1$, the set of shared variables $\vars(i)$ is non-empty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
However, there are also rules such as $\All{x~y}{P~x → Q~y → R~x~y}$, where slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
In this case, any match at level 1 can be extended with any hypothesis at level 2, since the two slots are independent.

@@

Now we add $h₂ : a ≤ b$, which again unifies with both slots.
For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $H_{y}(2, b) = \{h₁\}$.
Hence, we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
For slot~2, we add $h₂$ to $H_{y}(2, a)$.
There are no compatible matches at level~1, so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.

@@

To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
Hence, the compatible hypotheses are
\[
  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
\]
This means the match cannot currently be extended.
We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
This yields the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1, h_3\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}}
      child {node {$b$}
        child {node {$\{h_3\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$a$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
      child {node {$b$}
        child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}
Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
\]
This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.

@@

With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs are syntactically equal except for the marked proof subexpressions.
This notion of equality is syntactic, so it can be checked quickly, and we can define an effective hash function that is compatible with it.
This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.

@@

\para{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence, we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.

@@

\para{Lazy Insertion.}
Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
We therefore delay insertions in two ways.
First, we observe that complete matches can only be generated once all slots have been filled.
Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found a potentially matching hypothesis for each slot.
Second, we observe that the variable indices are only queried for hypotheses at a slot $i > 1$ when a match is added at slot $i - 1$.
Hence, we delay the insertion of hypotheses at $i$ until there is at least one match in slot $i - 1$.

@@

For each benchmark, we set up certain forward rules and a goal with hand-picked hypotheses.
We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
Performing only forward reasoning is the best-case scenario for the incremental algorithm since previous results can be fully reused.
When forward reasoning is combined with other rules, we would expect smaller gains, depending on how much reuse the non-forward rules allow.

@@

\newpage
\bibliography{lit}
